---
title: "Why net farm income is an incomplete yardstick for the farm safety net"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Why net farm income is an incomplete yardstick for the farm safety net}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## The question this package answers

How well is the farm safety net working? In U.S. policy discussion the de facto
answer is **annual net farm income (NFI)**, the headline sector aggregate that
USDA's Economic Research Service forecasts and that commentators track to judge
the "health" of the farm economy and, by extension, the programs that support it
(USDA-ERS, n.d.-a). Yet the stated purpose of those programs is not to raise the
average so much as to **manage risk**, that is, to dampen the year-to-year
variability of farm outcomes (Glauber, 2004; Coble & Barnett, 2013; Hardaker,
Lien, Anderson, & Huirne, 2015).

NFI is indispensable for what it was designed to do: summarize the sector's
income in a year. But as a *leading indicator for the safety net* it is
**incomplete**, because it does not measure what the safety net is for, and as a
single aggregate it obscures how outcomes are distributed across heterogeneous
farms (Mishra, El-Osta, Morehart, Johnson, & Hopkins, 2002). This vignette lays
out the complementary indicator that fills that gap. It draws on a connected
research program: farm-level evaluation of risk-reducing instruments (Tsiboe,
Tack, & Yu, 2023); large-scale calibration of the farm-level yields needed to do
so (Tsiboe, Turner, & Yu, 2025); the risk-reduction metrics themselves (Tsiboe et
al., 2025); the evaluation of program *combinations* (Gaku & Tsiboe, 2025); and
prospective assessment of proposed instruments (Tsiboe & Turner, 2025). The
final section shows how `rfsnEfficiency` operationalizes it.

## What is a farm safety-net instrument?

A **farm safety-net instrument** is any program or contract that moves income to
a producer in adverse states of the world, conditioned on a yield loss, a price
or revenue shortfall, or a disaster, in order to **stabilize farm outcomes**
(Glauber, 2004; Hardaker et al., 2015). Operationally, an instrument is defined
by four primitives:

1. a **trigger**: the adverse event that activates a payment (a farm yield below
   a guarantee, a county revenue below a benchmark, a declared disaster);
2. an **indemnity / payment rule**: how the payment maps to the shortfall;
3. a **producer cost**: a (often subsidized) premium or fee, or none; and
4. a **risk scope**: the pool over which losses are recognized, namely
   farm-level, county / area, or national.

The scope primitive is consequential, not cosmetic. Area- and index-based
triggers economize on information but introduce **basis risk**, the wedge
between the index and the farm's own outcome, which can sharply limit (or even
reverse) an instrument's farm-level risk reduction (Tsiboe et al., 2023). These
four primitives are exactly the inputs `rfsnEfficiency` takes (outcome draws,
plus an indemnity rule that may be a function, plus `cost` / `premium` /
`subsidy`), so any instrument, existing or proposed, can be scored on the same
footing.

## The U.S. farm safety net: current scope

Using the United States as context, federal programs span three groups (see the
[USFarmSafetyNetLab](https://github.com/ftsiboe/USFarmSafetyNetLab); CRS, 2021;
Glauber, 2004):

1. **Farm-specific risk management**: the Federal Crop Insurance Program (FCIP)
   and the Noninsured Crop Disaster Assistance Program (NAP), administered by
   USDA's Risk Management Agency (USDA-RMA, n.d.). Producers pay (subsidized)
   premiums; triggers are farm- or area-level yield/revenue outcomes. FCIP is the
   largest instrument and is itself layered (basic yield and revenue plans, unit
   structures, coverage levels, options, and supplemental endorsements such as
   SCO and ECO), and its premium subsidies are large enough that participation
   and design are shaped as much by the subsidy as by risk (Coble & Barnett,
   2013; Du, Feng, & Hennessy, 2017; ARPC, n.d.).
2. **Regional or national risk coverage**: Price Loss Coverage (PLC) and
   Agriculture Risk Coverage (ARC) under Title I, alongside marketing-loan and
   standing disaster programs. Producers pay no premium; triggers are national
   or county price/revenue benchmarks.
3. **Ad-hoc disaster assistance**: temporary, premium-free programs responding to
   events not well covered by the first two groups, for example pandemic- or
   trade-related assistance and large disaster packages.

Because a producer typically holds a **combination** across these groups (say, a
revenue FCIP policy plus PLC and, in a shock year, ad-hoc aid), the safety net's
performance is a property of the *combination*, not of any single instrument.
That is the central insight of combination-strategy evaluation (Gaku & Tsiboe,
2025). The menu is also not fixed: proposed instruments, such as a buy-up tier
for PLC, can be assessed *before* enactment for the risk reduction they would
deliver and at what cost (Tsiboe & Turner, 2025). Either way, evaluation requires
scoring configurable instruments on the same realized draws.

## Why NFI is an incomplete yardstick for the safety net

A risk-management program exists to **reduce the variability** of farm outcomes,
to convert volatile, loss-prone years into something steadier, not merely to
raise the average (Hardaker et al., 2015). Judging it by NFI alone leaves four
gaps:

1. **It targets the level, not the risk.** NFI tracks the mean. A program could
   leave mean income unchanged yet sharply cut its year-to-year swings, exactly
   the success a safety net is meant to deliver, and NFI would not move. Risk,
   moreover, is asymmetric: producers care chiefly about downside shortfalls
   below a target, which mean-level statistics ignore (Fishburn, 1977; Hardaker
   et al., 2015).
2. **It is an aggregate.** A single national figure hides the distribution,
   namely which crops, regions, and operations actually had their risk reduced
   and which did not, a long-standing limitation of sector-level income measures
   (Mishra et al., 2002). Farm-level evaluation is required to recover it
   (Tsiboe et al., 2023).
3. **It is partly circular.** NFI *includes* government payments (USDA-ERS,
   n.d.-b), so a more expensive program mechanically raises the very number used
   to judge it, regardless of whether it stabilized anything; this is sharpened
   by the fact that subsidies function in part as income transfers rather than
   pure risk protection (Coble & Barnett, 2013; Du et al., 2017).
4. **It is ex post and confounded.** Realized NFI mixes program effects with
   prices, weather, and macro conditions, so it cannot isolate the marginal
   contribution of a given configuration. Disentangling it requires a
   counterfactual, representative-agent comparison against a no-program baseline,
   which is feasible only with farm-level calibrated outcomes (Tsiboe, Turner, &
   Yu, 2025; Tsiboe et al., 2025).

## The alternative: efficacy and efficiency of risk reduction

Tsiboe et al. (2025) formalize the safety net's job as a comparison between the
distribution of a farm outcome **with** the program and the counterfactual
**without** it (the no-program baseline). Two questions follow.

**Efficacy, or does it work?** Two indices, each relative to the no-program
baseline (so `1` = no effect):

- **Income Transfer Score (ITS)**: the ratio of mean outcome with vs. without
  the program. `ITS > 1` means the program lifts the mean.
- **Variability Reduction Score (VRS)**: the ratio of an outcome-variability
  measure with vs. without the program. `VRS < 1` means it cuts variability.
  Because downside risk is what producers price, the package reports VRS not only
  on the coefficient of variation but also on loss-side (below-mean) partial
  moments (Fishburn, 1977).

A configuration "reduces risk" only when it does **both**: `ITS > 1` *and*
`VRS < 1`.

**Efficiency, or is it worth it?** The **Risk-Reduction Efficiency Ratio
(RRER)**, Equation (3) of Tsiboe et al. (2025):

$$
\text{RRER} \;=\; \mathbb{I}(\text{VRS}<1)\,\mathbb{1}(\text{ITS}>1)\;
\frac{1-\text{VRS}}{\text{ITS}-1}.
$$

RRER is the **variability reduction achieved per unit of income (outcome)
transferred**. It deflates the stabilization a program delivers by how much it
had to move the mean to get there. A program that simply hands out money every
year raises the mean a lot for little genuine stabilization and earns a *low*
RRER; a program that converts each transferred dollar into a large reduction in
volatility earns a *high* one. The same per-dollar logic underlies the
cost-effectiveness comparison of a proposed PLC buy-up tier in Tsiboe & Turner
(2025). Crucially, RRER is defined only when the configuration is already
efficacious: efficiency presumes efficacy.

## Keeping the indices well-posed: economic bounds

ITS, VRS, and RRER are all ratios, so an agent whose no-program baseline has a
mean or variability near zero can push an index to an implausible magnitude that
reflects a vanishing denominator rather than any economic effect. Across a
Monte-Carlo of $\sim\!10^{5}$ heterogeneous agents, a handful of such degenerate
cases would otherwise dominate the weighted or trimmed aggregates. To keep the
indicators well-posed, `rfsnEfficiency` winsorizes each index into an
economically feasible envelope (arguments `its_clamp`, `vrs_clamp`, and the RRER
`deadband` / `cap`), pulling the degenerate agents to the edge of the plausible
range instead of letting them set the headline. The defaults, and the economic
priors behind them, are:

| Index | Bound | Default | Economic reason |
|---|---|---|---|
| ITS (`mean_index`) | floor | 0.10 | Subsidized insurance cannot collapse mean revenue |
| ITS (`mean_index`) | cap | 2.00 | Transfers are a modest share of revenue; the mean cannot double |
| VRS (`cv_index`, downside, shortfall) | floor | 0.10 | Partial coverage and basis risk leave irreducible residual risk |
| VRS (`cv_index`, downside, shortfall) | cap | 2.00 | Insurance may add noise, but not multiply variability |
| RRER (`efficiency`) | floor | 0.00 | Zero unless the program both lifts the mean and cuts risk (definitional) |
| RRER (`efficiency`) | max | $1/\text{deadband}\approx 18\text{–}20$ | Efficiency is undefined near a zero income transfer |

Index by index:

1. **ITS floor (0.10).** Participation in a subsidized program is rational only
   when the expected net benefit is non-negative, so premium subsidies roughly
   offset the loading and $\mathbb{E}[\text{indemnity}+\text{subsidy}]\ge$ the
   producer premium (Coble & Barnett, 2013; Du et al., 2017). A functioning
   program therefore lifts, or at worst barely dents, mean revenue; an ITS of
   0.10 would mean it destroyed 90% of expected revenue, which can only arise
   from a baseline mean near zero.
2. **ITS cap (2.00).** Premiums are single-digit percentages of liability and
   subsidies cover roughly 60% of premium, so the mean uplift is on the order of
   a few percent. Because indemnities are bounded by liability
   ($\approx$ coverage $\times$ expected revenue), the mean simply cannot double;
   $\text{ITS}\gg 1.5$ signals a micro-baseline artifact.
3. **VRS floor (0.10).** Coverage tops out at $\le 90\%$, deductibles remain, and
   area endorsements (SCO/ECO) carry basis risk between the county pool and the
   farm (Tsiboe et al., 2023). That irreducible residual means variability cannot
   be driven to zero; the floor ("at most $\sim\!90\%$ of variability removed")
   is consistent with partial coverage and basis risk.
4. **VRS cap (2.00).** Insurance can add variance in small doses (premium outflow
   in good years; area basis risk), so mild $\text{VRS}>1$ is admitted, but
   doubling revenue variability is already extreme and larger values are
   tiny-baseline artifacts.
5. **RRER floor (0).** By Equation (3) the score is zero unless the configuration
   both lifts the mean and cuts variability, so non-negativity is built in, not
   imposed.
6. **RRER deadband (0.05).** RRER divides the proportional risk reduction by the
   proportional income gain, so as the mean gain $\text{ITS}-1\to 0$ the ratio
   explodes: a program that stabilizes with essentially no transfer looks
   infinitely efficient. Economically, a mean gain below $\sim\!5\%$ is within the
   noise of the transfer mechanism and is not a meaningful cost basis to divide
   by; requiring $\text{ITS}-1\ge 0.05$ before rating efficiency removes that
   singularity and mechanically bounds the score at $1/0.05=20$ (about 18 given
   the VRS floor).

These guardrails complement, rather than replace, the robust aggregation the
package already uses: the clamps fix per-agent artifacts at the source, trimmed
means with medians reported alongside handle the group aggregation, and the ITS
deadband removes the one remaining singularity (RRER as $\text{ITS}\to 1$). All
are optional; set the argument to `NULL` (or `deadband = 0`) to recover the raw
Equation (3) behaviour.

## What risk-reduction efficiency adds

| Property | Annual NFI | Risk-Reduction Efficiency |
|---|---|---|
| Measures risk (variability) | No (level only) | **Yes** |
| Captures downside / shortfall risk | No | **Yes** (Fishburn, 1977) |
| Distributional (by crop/region/agent) | No (aggregate) | **Yes** (Tsiboe et al., 2023) |
| Isolates the program's marginal effect | No (confounded) | **Yes** (vs. baseline) |
| Compares alternative configurations | No | **Yes** (Gaku & Tsiboe, 2025) |
| Independent of program spending | No (includes payments) | **Yes** (spending is the denominator, by design) |
| Forward / counterfactual capable | No (ex post) | **Yes** (Tsiboe & Turner, 2025) |

In short, RRER reports **how much risk reduction the safety net buys per dollar
it moves**, agent by agent. That is precisely the question a *risk-management*
program should be judged on, and precisely the dimension NFI leaves out. It does
not displace NFI; it completes the picture.

## How `rfsnEfficiency` operationalizes it

The package is program-agnostic: you supply each agent's end-of-season outcome
draws and the program parameters (an indemnity rule that may be a function, plus
cost/premium/subsidy), and it returns the scores and regime comparisons. The
outcome draws can come from calibrated farm-level yields of the kind developed in
Tsiboe, Turner, & Yu (2025).

```{r}
library(rfsnEfficiency)
data(toy_outcomes)   # built from the calibrated_yield release; see data-raw/

# Score a configuration and compare two regimes against the same draws
res <- run_efficiency_analysis(
  data = toy_outcomes, by = "group_id",
  outcome_args = list(base_value = "outcome", transfers = "indemnity",
                      premium = "premium"),
  ref = "baseline", alt = "reform")

res$scores   # ITS (mean_index), VRS (risk_index), RRER (efficiency) per regime
res$deltas   # reform - baseline for every score
```

Because the **outcome is configurable** (revenue, profit, margin, or a cost-side
measure), the same indicator can be reported on whichever farm outcome a policy
audience cares about, without changing the metric.

## What NFI still does well

This is not an argument to discard NFI. As a macro accounting aggregate it
remains the right tool for tracking sector income and budget exposure (USDA-ERS,
n.d.-a). The claim is narrower: **NFI is an incomplete yardstick for the safety
net, and a risk-reduction-efficiency indicator supplies the missing piece**, a
direct, distributional measure of the safety net's core function, reducing
farm-level risk efficiently. It can be computed retrospectively as a report card
and prospectively to compare reforms before they are enacted, on the same
realized shocks (Gaku & Tsiboe, 2025; Tsiboe & Turner, 2025; Tsiboe et al.,
2025).

## References

Agricultural Risk Policy Center (ARPC). (n.d.). *Federal Crop Insurance Program
backgrounders and policy reports.* North Dakota State University.
<https://github.com/ftsiboe/arpcFCIPHarness>

Coble, K. H., & Barnett, B. J. (2013). Why do we subsidize crop insurance?
*American Journal of Agricultural Economics, 95*(2), 498–504.
<https://doi.org/10.1093/ajae/aas093>

Congressional Research Service (CRS). (2021). *Federal crop insurance: A primer*
(Report R46686). <https://www.congress.gov/crs-product/R46686>

Du, X., Feng, H., & Hennessy, D. A. (2017). Rationality of choices in subsidized
crop insurance markets. *American Journal of Agricultural Economics, 99*(3),
732–756. <https://doi.org/10.1093/ajae/aaw035>

Fishburn, P. C. (1977). Mean-risk analysis with risk associated with below-target
returns. *American Economic Review, 67*(2), 116–126.

Gaku, S., & Tsiboe, F. (2025). Evaluation of alternative farm safety net program
combination strategies. *Agricultural Finance Review, 85*(2), 254–273.
<https://doi.org/10.1108/AFR-11-2023-0150>

Glauber, J. W. (2004). Crop insurance reconsidered. *American Journal of
Agricultural Economics, 86*(5), 1179–1195.
<https://doi.org/10.1111/j.0002-9092.2004.00663.x>

Hardaker, J. B., Lien, G., Anderson, J. R., & Huirne, R. B. M. (2015). *Coping
with risk in agriculture: Applied decision analysis* (3rd ed.). CABI.

Mishra, A. K., El-Osta, H. S., Morehart, M. J., Johnson, J. D., & Hopkins, J. W.
(2002). *Income, wealth, and the economic well-being of farm households*
(Agricultural Economic Report No. 812). USDA Economic Research Service.

Tsiboe, F., Tack, J., & Yu, J. (2023). Farm-level evaluation of area- and
agroclimatic-based index insurance. *Journal of the Agricultural and Applied
Economics Association, 2*(4), 616–633. <https://doi.org/10.1002/jaa2.77>

Tsiboe, F., Turner, D., & Yu, J. (2025). Utilizing large-scale insurance data
sets to calibrate sub-county level crop yields. *Journal of Risk and Insurance,
92*(1), 139–165. <https://doi.org/10.1111/jori.12494>

Tsiboe, F., Turner, D., Williams, B., Miller, M., Baldwin, K., & Dohlman, E.
(2025). Risk reduction impacts of crop insurance in the United States. *Applied
Economic Perspectives and Policy, 47*(5), 1832–1847.
<https://doi.org/10.1002/aepp.13513>

Tsiboe, F., & Turner, D. (2025). Incorporating buy-up price loss coverage into
the United States farm safety net. *Applied Economic Perspectives and Policy,
47.* <https://doi.org/10.1002/aepp.13536>

U.S. Department of Agriculture, Economic Research Service (USDA-ERS). (n.d.-a).
*Farm sector income and finances.*
<https://www.ers.usda.gov/topics/farm-economy/farm-sector-income-finances>

U.S. Department of Agriculture, Economic Research Service (USDA-ERS). (n.d.-b).
*Farm income and wealth statistics.*
<https://www.ers.usda.gov/data-products/farm-income-and-wealth-statistics>

U.S. Department of Agriculture, Risk Management Agency (USDA-RMA). (n.d.).
*Summary of Business and program information.* <https://www.rma.usda.gov>
